---
title: "Persulfidation: a Proteomic Analysis"
author: "Bryan Montejano Aldama"
date: "2024-12-13"
output: 
  pdf_document:
    toc: true         
    number_sections: true 
fontsize: 10pt        
geometry: a4paper      
header-includes:       
  - \usepackage{setspace}
  - \setstretch{1.5}   
  - \usepackage{float} 
---

# Anexo III
```{r setup, include=FALSE, message=FALSE}
setwd("/home/nicolasdi/BIXO-LINUX/-Persulfidation--A-proteomic-Analysis")

required_packages <- c("VennDiagram", "grid", "futile.logger", "ggplot2", "dplyr", "pacman", "readr", 
                       "BiocManager", "AnnotationDbi", "org.Rn.eg.db", "clusterProfiler", "factoextra", "biomaRt")

# Install missing CRAN packages
for (pkg in required_packages) {
  if (pkg %in% c("BiocManager", "AnnotationDbi", "org.Rn.eg.db", "clusterProfiler", "biomaRt")) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      BiocManager::install(pkg)
    }
  } else {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg)
    }
  }
}

library(pacman)
# Load all necessary libraries at once using pacman
pacman::p_load(VennDiagram, grid, futile.logger, ggplot2, dplyr, factoextra, clusterProfiler, 
               org.Rn.eg.db, AnnotationDbi, biomaRt, fastmap, readr, dplyr, tidyr, gapminder, tidyverse, cowplot)

```
## Descriptive Analysis

### Coefficient of Variation

The coefficient of variation was calculated as the ratio of the standard deviation to the mean of the data, and expressed as a percentage: $$CV= (Standard \ deviation /mean)*100$$ For this and the following analyses, the file "~/Proteins/b.csv" was used. 

```{r, warning=FALSE, message=FALSE}
# Set the directory name
base_dir <- "/home/nicolasdi/BIXO-LINUX/-Persulfidation--A-proteomic-Analysis/" 

file_path <- paste0(base_dir, "/Proteins/b.csv")

# Load the .csv file used 
a <- read_csv(file_path)
a <- data.frame(a, row.names = c("C1", 
                                 "C2", 
                                 "C3", 
                                 "O1", 
                                 "O2", 
                                 "O3", 
                                 "S1", 
                                 "S2", 
                                 "S3")) 
proteins <- data.frame(colnames(a)) # Extracted the names of the 1395 proteins

# From the three experimental conditions: Control ("C"), NaSH ("S"), and OGD ("O"), the CV function was applied to each protein using the relative abundance values. Three figures were obtained, showing the CV of each protein relative to the log10 of its mean.

# Function to calculate the mean and CV for a given subset of rows
calculate_cv_mean <- function(data, rows) {
  mean_vals <- sapply(data[rows, ], mean)  # Calculate mean for the selected rows
  cv_vals <- sapply(data[rows, ], function(x) sd(x) / mean(x) * 100)  # Calculate CV for the selected rows
  return(data.frame(mean = mean_vals, cv = cv_vals))  # Return as a data frame
}

# Control (CV and mean for Control samples)
dfC <- calculate_cv_mean(a, 1:3)

# NaHS (CV and mean for NaHS samples)
dfS <- calculate_cv_mean(a, 7:9)

# OGD (CV and mean for OGD samples)
dfO <- calculate_cv_mean(a, 4:6)


# Calculate log10 of the mean for each data frame explicitly
dfC <- mutate(dfC, log_x = log10(dfC$mean))  # Apply log10 to the 'mean' column in dfC
dfS <- mutate(dfS, log_x = log10(dfS$mean))  # Apply log10 to the 'mean' column in dfS
dfO <- mutate(dfO, log_x = log10(dfO$mean))  # Apply log10 to the 'mean' column in dfO


combined_data <- bind_rows(
  mutate(dfC, Muestra = "Control"),
  mutate(dfS, Muestra = "NaHS"),
  mutate(dfO, Muestra = "OGD")
)

  # Plot for all samples
ggplot(combined_data, aes(x = log_x, y = cv, color = Muestra)) +
  geom_point(size = 1) +
  ggtitle("Coefficient of Variation by Sample Type") +
  xlab("log10(Mean)") + ylab("CV (%)") +
  scale_color_manual(values = c("Control" = "blue", "NaHS" = "red", "OGD" = "green")) +
  theme_minimal()
  
```
------------------------------------------------------------------------

## PCA

The PCA analysis represents the variation in the dataset expressed in transformed variables (Principal Components). The factorextra package is used for PCA. Data was loaded from the file "~/Proteins/b.csv" and processed as input.
Data standardization was performed using a robust approach based on the median and interquartile range (IQR) for the 1395 variables (proteins) using the formula:

$$ Robust\ standardization= (X -mean)/IQR $$

```{r, include=T, warning=FALSE, message=FALSE}
# Calculate median and IQR for each protein (across 9 samples)
mediana <- apply(a, 2, median)  # Median for each column (protein)
iqr <- apply(a, 2, IQR)         # IQR for each column (protein)

#Function for standardization
robust_scale <- function(x, mediana, iqr)
  { y=(x-mediana)/(iqr)} 

#Applying robust_scale function for each protein
res <- data.frame (C1 = c(), 
                   C2 = c(), 
                   C3 = c(), 
                   O1 = c(), 
                   O2 = c(), 
                   O3 = c(), 
                   S1 = c(), 
                   S2 = c(), 
                   S3 = c()) 
  
for (j in 1:1395) {
  for (i in 1:9){
    res[i,j] <- robust_scale(a[i,j], mediana[j], iqr[j])}
  }

# Transpose the result and remove NA values
res <- data.frame(res, row.names = c("C1", 
                                     "C2", 
                                     "C3", 
                                     "O1", 
                                     "O2", 
                                     "O3", 
                                     "S1", 
                                     "S2", 
                                     "S3")) 

res <- data.frame(t(res)) #Results "res" is structured for further analysis
res <- data.frame (res, row.names = t(proteins)) 
res <- data.frame(t(res)) 

res1 <- data.frame(t(na.omit(t(res)))) 

# PCA is performed
pca <- prcomp(res1, scale = F) 

#Explained variation can be seen with summary
summary(pca) 

# Visualitazion
pcaind <-as.data.frame(pca$x) # Names were extracted

ggplot(data = pcaind, mapping = aes(y=PC1, x=PC2, z=PC3)) +
geom_point(color=c('red', 'red', 'red', 'green', 'green', 'green', 'blue', 'blue', 'blue')) +
  ggtitle("Análisis de componentes principales")+
  xlab("PC1 (90.5%)")+ #Obtenido con la variación explicada 
  ylab("PC2(6.2%)")+
  geom_text(label=c("C1", "C2", "C3", "O1", "O2", "O3", "S1", "S2", "S3"))
```

## Persulfurated protein identification

Proteins with the label identified as **DCP-Bio1** or **DCP-Bio1 hydrolyzed** were filtered from the .csv files generated in the qualitative analysis. The code for the control condition is explained in detail in 4 steps. The same procedure was applied to the other two conditions.

```{r message=FALSE, warning=FALSE, include=T}

# Step 1: The .csv file of each of the 3 biological replicates was loaded as a data frame. Proteins were filtered by those containing "DCP-Bio1" or "DCP-Bio1 hydrolyzed" in the "xxx" column for each replicate.

# Data loading

C1 <- read_csv(paste0(base_dir, "/Proteins/C1.csv"))
C2 <- read_csv(paste0(base_dir, "/Proteins/C2.csv"))
C3 <- read_csv(paste0(base_dir, "/Proteins/C3.csv"))

# Step 2: Data filtering for those with "DCP-Bio1" in the "peptide.modification" column and reducing the data frames to 3 columns: "protein.Accession", "protein.Entry.x", and "protein.Description"

C1filtered <- C1[grepl("DCP-Bio1", C1$peptide.modification), ]
C1filtered <- data.frame(C1filtered[, 2:4])

C2filtered <- C2[grepl("DCP-Bio1", C2$peptide.modification), ]
C2filtered <- data.frame(C2filtered[, 2:4])

C3filtered <- C3[grepl("DCP-Bio1", C3$peptide.modification), ]
C3filtered <- data.frame(C3filtered[, 2:4])

# Step 3: The 3 replicates were consolidated into a single data frame. To better identify the proteins, the "split" function was used on the "protein.Description" column to split the text into columns.

C1C2C3 <- rbind(C1filtered, C2filtered, C3filtered)

C1C2C3T <- data.frame(names = C1C2C3$protein.Description)

protein_list_Control <- C1


C1C2C3T <- C1C2C3T %>%
  mutate(split = strsplit(as.character(names), "=")) %>%
  transmute(
    Persulfurated_Proteins = sapply(split, `[`, 1),
    key2 = sapply(split, `[`, 2),
    key3 = sapply(split, `[`, 3),
    key4 = sapply(split, `[`, 4),
    key5 = sapply(split, `[`, 5),
    key6 = sapply(split, `[`, 6)
  )

# Step 4: A list was generated indicating the identification of persulfurated proteins.

PersulfuratedControl <- C1C2C3T %>% distinct(Persulfurated_Proteins)
PersulfuratedControl_protein.accession <- C1C2C3 %>% distinct(C1C2C3$protein.Accession)
PersulfuratedControl_protein.accession <- as.character(PersulfuratedControl_protein.accession$`C1C2C3$protein.Accession`)

## NaHS 
# Step 1
S1 <- read_csv(paste0(base_dir, "/Proteins/S1.csv"))
S2 <- read_csv(paste0(base_dir, "/Proteins/S2.csv"))
S3 <- read_csv(paste0(base_dir, "/Proteins/S3.csv"))

# Step 2

S1filtered <- S1[grepl("DCP-Bio1", S1$peptide.modification), ]
S1filtered <- data.frame(S1filtered[, 2:4])

S2filtered <- S2[grepl("DCP-Bio1", S2$peptide.modification), ]
S2filtered <- data.frame(S2filtered[, 2:4])

S3filtered <- S3[grepl("DCP-Bio1", S3$peptide.modification), ]
S3filtered <- data.frame(S3filtered[, 2:4])

# Step 3

S1S2S3 <- rbind(S1filtered, S2filtered, S3filtered)

S1S2S3T <- data.frame(names = S1S2S3$protein.Description)

library(tidyr)
S1S2S3T <- S1S2S3T %>%
  mutate(split = strsplit(as.character(names), "=")) %>%
  transmute(
    Persulfurated_Proteins = sapply(split, `[`, 1),
    key2 = sapply(split, `[`, 2),
    key3 = sapply(split, `[`, 3),
    key4 = sapply(split, `[`, 4),
    key5 = sapply(split, `[`, 5),
    key6 = sapply(split, `[`, 6)
  )

# Step 4

PersulfuratedNaHS <- S1S2S3T %>% distinct(Persulfurated_Proteins)
PersulfuratedNaHS_protein.accession <- S1S2S3 %>% distinct(S1S2S3$protein.Accession)
PersulfuratedNaHS_protein.accession <- as.character(PersulfuratedNaHS_protein.accession$`S1S2S3$protein.Accession`)

# OGD 

# Step 1
O1 <- read_csv(paste0(base_dir, "/Proteins/O1.csv"))
O2 <- read_csv(paste0(base_dir, "/Proteins/O2.csv"))
O3 <- read_csv(paste0(base_dir, "/Proteins/O3.csv"))

# Step 2

O1filtered <- O1[grepl("DCP-Bio1", O1$peptide.modification), ]
O1filtered <- data.frame(O1filtered[, 2:4])

O2filtered <- O2[grepl("DCP-Bio1", O2$peptide.modification), ]
O2filtered <- data.frame(O2filtered[, 2:4])

O3filtered <- O3[grepl("DCP-Bio1", O3$peptide.modification), ]
O3filtered <- data.frame(O3filtered[, 2:4])

# Step 3

O1O2O3 <- rbind(O1filtered, O3filtered, O2filtered)

O1O2O3T <- data.frame(names = O1O2O3$protein.Description)

library(tidyr)
O1O2O3T <- O1O2O3T %>%
  mutate(split = strsplit(as.character(names), "=")) %>%
  transmute(
    Persulfurated_Proteins = sapply(split, `[`, 1),
    key2 = sapply(split, `[`, 2),
    key3 = sapply(split, `[`, 3),
    key4 = sapply(split, `[`, 4),
    key5 = sapply(split, `[`, 5),
    key6 = sapply(split, `[`, 6)
  )

# Step 4

PersulfuratedOGD <- O1O2O3T %>% distinct(Persulfurated_Proteins)
PersulfuratedOGD_protein.accession <- O1O2O3 %>% distinct(O1O2O3$protein.Accession)
PersulfuratedOGD_protein.accession <- as.character(PersulfuratedOGD_protein.accession$`O1O2O3$protein.Accession`)
```

### Protein visualization

```{r, include=T, warning=FALSE, message=FALSE}
# Transformation of data frames into vectors
PersulfuratedControl <- as.character(PersulfuratedControl$Persulfurated_Proteins)
PersulfuratedNaHS <- as.character(PersulfuratedNaHS$Persulfurated_Proteins)
PersulfuratedOGD <- as.character(PersulfuratedOGD$Persulfurated_Proteins)

# Creation of the Venn diagram
venn.plot <- venn.diagram(
    x = list(Set1 = PersulfuratedControl, 
             Set2 = PersulfuratedNaHS, 
             Set3 = PersulfuratedOGD),
    category.names = c("Control", "NaHS", "OGD"),
    fill = c("red", "blue", "green"),  # Circle colors
    alpha = 0.5,  # Transparency of the colors
    cat.cex = 1.5,  # Category text size
    cex = 1.5,  # Number text size
    fontface = "bold",  # Text style
    cat.fontface = "bold",
    filename = NULL)  

# Draw the diagram
grid.draw(venn.plot)
```

#### Control
```{r, include=T, warning=FALSE, message=FALSE}
# Proteins identified only in the Control condition
Venn_Control <- setdiff(PersulfuratedControl, union(PersulfuratedNaHS, PersulfuratedOGD))
Lista_Control <- setdiff(PersulfuratedControl_protein.accession, union(PersulfuratedNaHS_protein.accession, PersulfuratedOGD_protein.accession))
knitr::kable(sort(Venn_Control))
```

#### Control and NaHS
```{r, include=T, warning=FALSE, message=FALSE}
# Proteins identified only in the Control and NaHS condition
Venn_Control_X_NaHS <- setdiff(intersect(PersulfuratedControl, PersulfuratedNaHS), PersulfuratedOGD)
Lista_Control_X_NaHS <- setdiff(intersect(PersulfuratedControl_protein.accession, PersulfuratedNaHS_protein.accession), PersulfuratedOGD_protein.accession)
knitr::kable(Venn_Control_X_NaHS)
```

#### Control and OGD
```{r, include=T, warning=FALSE, message=FALSE}
# Proteins identified only in the Control and OGD condition
Venn_Control_X_OGD <- setdiff(intersect(PersulfuratedControl, PersulfuratedOGD), PersulfuratedNaHS)
Lista_Control_X_OGD <- setdiff(intersect(PersulfuratedControl_protein.accession, PersulfuratedOGD_protein.accession), PersulfuratedNaHS_protein.accession)
knitr::kable(sort(Venn_Control_X_OGD))
```

#### Control, NaHS y OGD
```{r, include=T, warning=FALSE, message=FALSE}
# Proteins identified only in the Control, NaHS, and OGD condition
Venn_Control_X_OGD_X_NaHS <- Reduce(intersect, list(PersulfuratedControl, PersulfuratedNaHS, PersulfuratedOGD))
Lista_Control_X_OGD_NaHS <- Reduce(intersect, list(PersulfuratedControl_protein.accession, PersulfuratedNaHS_protein.accession, PersulfuratedOGD_protein.accession))
knitr::kable(sort(Venn_Control_X_OGD_X_NaHS))
```

#### NaHS
```{r, include=T, warning=FALSE, message=FALSE}
# Proteins identified only in the NaHS condition
Venn_NaHS <- setdiff(PersulfuratedNaHS, union(PersulfuratedControl, PersulfuratedOGD))
Lista_NaHS <- setdiff(PersulfuratedNaHS_protein.accession, union(PersulfuratedControl_protein.accession, PersulfuratedOGD_protein.accession))
knitr::kable(sort(Venn_NaHS))
```

#### OGD
```{r, include=T, warning=FALSE, message=FALSE}
# Proteins identified only in the OGD condition
Venn_OGD <- setdiff(PersulfuratedOGD, union(PersulfuratedControl, PersulfuratedNaHS))
Lista_OGD <- setdiff(PersulfuratedOGD_protein.accession, union(PersulfuratedControl_protein.accession, PersulfuratedNaHS_protein.accession))
knitr::kable(sort(Venn_OGD))
```

#### OGD y NaHS
```{r, include=T, warning=FALSE, message=FALSE}
# Proteins identified only in the OGD and NaHS conditions
Venn_OGD_X_NaHS <- setdiff(intersect(PersulfuratedOGD, PersulfuratedNaHS), PersulfuratedControl)
Lista_OGD_X_NaHS <- setdiff(intersect(PersulfuratedOGD_protein.accession, PersulfuratedNaHS_protein.accession), PersulfuratedControl_protein.accession)
knitr::kable(sort(Venn_OGD_X_NaHS))
```

## Análisis Gene Ontology (GO)

El análisis GO comenzó con la extracción del identificador Uniprot vinculado a la identificación del péptido con la señal de DCP-Bio. Luego, este identificador se tradujo a ENSEMBL utilizando la biblioteca "biomaRt". Para cada condición, se llevó a cabo el análisis GO utilizando herramientas como "clusterProfiler", "AnnotationDbi" y la base de datos de Rattus norvegicus como fondo ("org.Rn.eg.db").

```{r, include=T, warning=FALSE, message=FALSE}
# Function to perform GO enrichment and generate barplot
enrich_and_plot <- function(gene_list, org_db, key_type, ontology, show_category, axis_text_size = 8) {
  # Perform GO Enrichment
  enrichment_result <- enrichGO(
    gene = gene_list,
    OrgDb = org_db,
    keyType = key_type,
    ont = ontology
  )
  
  # Generate barplot with axis text customization
  plot <- barplot(enrichment_result, showCategory = show_category) +
    theme(
      axis.text.x = element_text(size = axis_text_size, angle = 45, hjust = 1),
      axis.text.y = element_text(size = axis_text_size),
      axis.title = element_text(size = axis_text_size + 2),
      plot.title = element_text(size = axis_text_size + 4, face = "bold"),
      legend.text = element_text(size = axis_text_size + 1)
    )
  
  return(plot)
}

# Function to save plot as PNG
save_plot <- function(plot, output_file, res = 250, width = 1400, height = 1800) {
  png(output_file, res = res, width = width, height = height)
  print(plot)
  dev.off()
}

# Define global parameters
org_db <- "org.Rn.eg.db"
key_type <- "UNIPROT"
ontology <- "BP"
show_category <- 15
axis_text_size <- 8

# Define gene lists
gene_lists <- list(
  Control = Lista_Control,
  Control_NaHS = Lista_Control_X_NaHS,
  Control_OGD = Lista_Control_X_OGD,
  Control_OGD_NaHS = Lista_Control_X_OGD_NaHS,
  NaHS = Lista_NaHS,
  OGD = Lista_OGD,
  OGD_NaHS = Lista_OGD_X_NaHS
)

# Directory to save plots
output_dir <- "plots/"  
if (!dir.exists(output_dir)) dir.create(output_dir)

# Process all gene lists and save plots
plots <- lapply(names(gene_lists), function(name) {
  cat("Processing:", name, "\n")

  # Generate plot
  plot <- enrich_and_plot(
    gene_list = gene_lists[[name]],
    org_db = org_db,
    key_type = key_type,
    ontology = ontology,
    show_category = show_category,
    axis_text_size = axis_text_size
  )
  
  # Save plot
  output_file <- paste0(output_dir, "GO_", name, "_Plot.png")
  save_plot(plot, output_file)
})

```

